Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-04-10 analysis/paper/paper.Rmd
Some more writing and introducing spatial data for map overview.

```{r libraries, eval = TRUE}
library(tidyverse)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(sf)
```

```{r get-data, eval = TRUE}
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))

@@

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [@reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016]. Some studies have attempted more comprehensive evaluations of entire lithic assemblages to get at potential site types and associated mobility patterns [@viken2018a; @breivik2020], but these have predominantly been conducted without support from quantitative frameworks that can help structure such analyses. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods. The analysis conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites. This is done to identify latent patterns and structure in the relationship between these, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency, and group sizes at visits to the sites. Furthermore, the assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.

@@

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas. Furthermore, these coastal sites appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the location of the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. The shore-bound settlement combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of shoreline displacement allows for a relatively good chronological control of these accumulation events. In other parts of the world, a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.

The `r nrow(assemblages)` sites chosen for analysis here have a relatively limited geographical distribution. The sites were excavated as part of four larger projects all that took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The institution responsible for all of these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the projects. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [e.g. @clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences. While certainly difficult to fully account for, I do believe that the relative contemporaneity of the projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis aimed at revealing behavioural rather than analytical patterning.

@@

Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data, without having to impose overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate should be of value to any subsequent in-depth studies of individual sites, and as a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in the literature, and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues.

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified in the original excavation reports. The purpose of this exercise is to evaluate the degree to which the assemblages follow a chronological structure as has been suggested by previous literature based on more informally driven methods. While some artefact categories will be combined in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful. However, given the vast inferential leap that this assumption involves, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with this limitation in mind. These measures are taken from the framework of whole assemblage behavioural index, as developed by , and is aimed to get at more specific behavioural dimensions relating to mobility patterns among hunter-gatherers. Through a series of studies, show that the relationship between lithic volumetric density and relative frequency of retouched artefacts provides consistent and comparable results across a wide range of chronological and cultural context. These measures have, however, to my knowledge not seen application in Scandinavian archaeology as of yet. Following the continuous operationalisations of these measures, principal components analysis (PCA) is more suitable than CA for this part of the analysis [e.g. @baxter1994, 100; @Bicho2020].

@@

lithics2 <- assemblages %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
  # The reason for not using .keep below is for the option to retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake, flake_other, 
                     fragment_other, chip_other, blade_other,  microblade_other, 
                     core_other, core_fragment_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count) %>% 
  column_to_rownames("site")
```

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020. While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the frequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here. If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility that sites from later phases. @Bicho2020 also found that WABI, as introduced by, gave a reasonable estimation of mobility in their data. The same is evident here where fig indicates a

@@

The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking. While this result is perhaps not surprising, given the previous indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for comparative studies. The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, while this can perhaps support the of the . The question then becomes what is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, another possibility is that these are indicative of site types that are temporally ubiquitous. In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020. This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to the elements, and failed to identify any diachronic variation in this pattern. As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similar, one could envisage that variablity in site functionality indicated by the core to chip correlation might be reflected in variation in degree of exposure.

>>>>

```{r libraries, eval = TRUE}
library(tidyverse)
library(patchwork)
library(gt)
library(FactoMineR)
library(factoextra)
library(sf)
library(rnaturalearth)
```

```{r get-data, eval = TRUE}
assemblages <- read.csv(here::here('analysis/data/raw_data/assemblage_data.csv'))
site_pts <- st_read('analysis/data/raw_data/sites.gpkg')
countries <- ne_download(scale = 110, type = 'land', category = 'physical',
                         returnclass = "sf")
```

@@

Previous studies of lithic artefacts in Norway have typically had a focus on chronological trends associated with formal tool types [e.g. @helskog1976; @reitan2016], involved refit-studies of the entirety of individual assemblages [@skar1986], or in-depth analyses concerned with technological processes associated with certain sub-categories of the site inventories [@damlien2016]. Some studies have attempted more comprehensive evaluations of entire lithic assemblages to get at potential site types and associated mobility patterns [@viken2018a; @breivik2020], but these have predominantly been conducted without support from quantitative frameworks that can help structure such analyses. In sum, therefore, previous studies are typically either limited to a small number of sites, to a subset of the inventories, or to narratively and subjectively driven methods. The analysis conducted here employs multivariate exploratory statistics using the entirety of the lithic assemblages associated with a larger number of sites. This is done to identify latent patterns and structure in the relationship between these, with the ultimate aim of identifying behaviourally induced variation in their composition across time and space. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency, and group sizes at visits to the sites. Furthermore, the assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern.

Although each individual assemblage can potentially have been impacted by an infinitude of effects that might skew an archaeological interpretation, this does not preclude the applicability of inductive analyses aimed at revealing overarching structure in the data without imposing overly complex analytical frameworks that attempt to account for these particularities [@bevan2015]. Structure that can be revealed from considering all of the assemblages in aggregate can constitute a step in an iterative analytical chain that ultimately aims to tease apart such aspects from the multitude of factors that have shaped the composition of the assemblages, and should be of value to any subsequent in-depth studies of individual sites. The most immediate danger of the approach outlined here is rather to be overly naive in the causal significance and cultural importance that is ascribed to any identified pattern. As such, the main aim of this analysis is to compare the results with findings reported in previous literature, and have the generation of new hypotheses as a possible outcome. To this end, the analysis follows two analytical avenues.

@@

The `r nrow(assemblages)` coastal sites chosen for analysis here have a relatively limited geographical distribution in south-eastern Norway. The sites were excavated as part of four larger excavation projects that all took place within the last 15 years [@solheim2013b; @jaksland2014a; @melvold2014b; @reitan2014l; @solheim2017b]. The institution responsible for all of these excavations was the Museum of Cultural History in Oslo. This has led to a considerable overlap in the archaeological personnel involved, and comparable excavation practices across the excavations. Furthermore, with these projects, major efforts were made to standardise how lithic artefacts were to be classified at the Museum of Cultural History. As a result, this should reduce the amount of artificial patterning in the data incurred by discrepancies in the employed systems for categorisation [@clark2006; @dibble2017]. In this setting, for example, bias could potentially follow from the fact that two of the projects have sites with relatively contemporaneous dates [@solheim2013b; @jaksland2014]. Any project-dependent classification practice could as a consequence lead to an exaggeration of chronological differences between the assemblages. While certainly difficult to fully account for, I do believe that the relative contemporaneity of the excavation projects, as well as the overlap in excavation and classification practices should minimise the above-mentioned effects, making the data-set a good candidate for exploratory data analysis.

A defining characteristic of the Norwegian Mesolithic is that a clear majority of the known sites are located in coastal areas. Furthermore, coastal sites, including those treated here, appear to predominantly have been located on or close to the contemporary shoreline when they were in use [@breivik2018; @solheim2020]. In south-eastern Norway, this pattern is combined with a continuous regression of the shoreline, following from isostatic rebound [e.g. @sørensen1979; @romundset2018]. The fairly rapid shoreline displacement means that the location of the sites tend not to have retained their strategic or ecologically beneficial shore-bound location for long periods of time [cf. @perreault2019, 47]. The shore-bound settlement, combined with the rapid shoreline displacement has resulted in a relatively high degree of spatial separation of cumulative palimpsests, to follow the terminology of @bailey2007, while the reconstruction of the trajectory of shoreline displacement combined with shore-bound settlement allows for a relatively good chronological control of these accumulation events. In other parts of the world a higher degree of spatial distribution means that while the physical separation of material can help delineate discrete events, this typically comes at the cost of loosing temporal resolution, as any stratigraphic relationship between the events is lost [@bailey2007]. However, as the rate of isostatic rebound has varied throughout the Mesolithic in the region, and local topography and bathymetry will have impacted how rapidly a site lost its shore-bound location, this effect is not evenly distributed in time and space. In the earliest part of the Mesolithic, the displacement rate within the study area would have been around as much as 8.8 cm/year, which is reduced to around 0.5 cm/year in the Late Mesolithic [@sørensen2014]. Thus, while relative sea-level change appears to have reduced the degree of mixing that has occurred in the assemblages, this could vary depending on when and where they were in use, in turn potentially reducing the degree to which their composition can be directly compared - a point that is returned to in the discussion of the results.

In a series of case studies from Northern Norway, @bølviken1982 were among a group of scholars often attributed the popularisaton of correspondence analysis (CA) in wider archaeologcal circles [@baxter1994]. In southern Norway, however, the use of multivariate statistics in Mesolithic research is to my knowledge limited to @solheim2013a use of CA to analyse the assemblage data from eight sites excavated in connection with a development-led project. As these sites, which are included in this study, have relatively contemporaneous dates, the Solheims analysis was mainly aimed at

There have been a few studies that have conducted comprehensive analyses of artefact inventories and site features to get at variation in Mesolithic site types [@viken2018; @breivik2020]. However, the narratively driven analyses of these data makes the weighting of different variables unclear and subjectively defined, and several of the studies cited here ultimately draw their conclusions based on an arguably over-constraining trichotomy of site types. As Binford originally clarified, which is a point that has also been retiterated by several authors since; these categories are best understood as extremes on a continuous scale.

@@

The first part of the analysis involves employing the method of correspondence analysis (CA), using the lithic count data as classified for the original excavation reports. The purpose of this exercise is to evaluate the degree to which the assemblages align with patterns that have been suggested by earlier application of more informally driven methods. The expectation is that chronological trends will be the main driver of While some artefact categories will be combined in an attempt to reduce the amount of subjective discretion that can be expected to have impacted the classification of the individual assemblages, this will largely involve an analysis that assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviorally meaningful.

However, given the vast inferential leap that this assumption involves, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages developed with this limitation in mind. These measures are taken from the framework of whole assemblage behavioural index, as developed by , and is aimed to get at more specific behavioural dimensions relating to mobility patterns among hunter-gatherers. Through a series of studies, show that the relationship between lithic volumetric density and relative frequency of retouched artefacts provides consistent and comparable results across a wide range of chronological and cultural context. These measures have, however, to my knowledge not seen application in Scandinavian archaeology as of yet. Following the continuous operationalisations of these measures, principal components analysis (PCA) is more suitable than CA for this part of the analysis [e.g. @baxter1994, 100; @Bicho2020].

@@

lithics2 <- sites %>% 
  select(-n, -beach_flint_flint_nodule, -hammerstone, -excavation_project) %>% 
  # The reason for not using .keep below is for the option to retain artefact_count
  mutate(chip_freq =  sum(chip, chip_other) / artefact_count,
         core_freq = sum(core, core_other) / artefact_count,
         primary_freq = sum(core, core_fragment, blade, 
                     microblade, chip, fragment, flake, flake_other, 
                     fragment_other, chip_other, blade_other,  microblade_other, 
                     core_other, core_fragment_other) / artefact_count,
         secondary_freq =  sum(axe_debitage, axe, chisel, 
                microlith, tanged_point, transverse_point, single_edged_point,
                hognipen_point, projectile_undefined, microburin, scraper, 
                knife, burin, drill, blade_retouch, microblade_retouch, 
                chip_retouch, fragment_retouch, flake_retouch, axe_other,
                axe_debitage_other, chisel_other, club_hatchet_other, 
                grinding_slab_stone, projectile_point_other, knife_other,
                scraper_other, drill_other, microblade_retouch_other, 
                blade_retouch_other, fragment_retouch_other,
                flake_retouch_other) / artefact_count,
         lithic_dens = artefact_count / excavated_m3) %>%
  select(-primary_lithics, -secondary_lithics, -excavated_m3, -artefact_count) %>% 
  column_to_rownames("site")
```

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for degree of mobility as operationalised by @Bicho2020. While the investigation performed by @Bicho2020 indicates that the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns than the volumetric density of lithics, and the frequency of retouched artefacts, these last dimensions are capturing the most variation in the dataset presented here. If frequency of secondarily worked artefacts is taken as a proxy for mobility, these findings would consequently be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases. @Bicho2020 also found that WABI, as introduced by, gave a reasonable estimation of mobility in their data. The same is evident here where fig indicates a

@@

The consistent nature of the negative correlation between LVD and relative retouch frequency across this study (r = -0.50), @clark2017 (r = -0.50) and @Bicho2020 (r = -0.48) is striking. While this result is perhaps not surprising given the previous and plentiful indications of the same trends, these results do speak to the cross-cultural and cross-temporal applicability of the measure, and consequently the potential it holds for larger scale comparative studies. The findings indicated by the WABI, as well as by the negative correlation between relative frequency of primarily and secondarily worked lithics, do align more clearly with previous suggestions concerning the overall mobility patterns in Mesolithic Norway. However, the question then becomes: What is the significance of the negatively correlated variables of relative frequency of cores and chips? Given that these are more or less orthogonal to the WABI variables on the first and second component, this would indicate that if one of these variable pairs is taken to reflect the of expedient and curated assemblages, it should follow that the other variable pairs is not correlated with these. As the entire temporal range of the sites is spread out between the extremes of the chip and core variables, one possibility is that these are indicative of site types that are temporally ubiquitous. In attempt to determine if this might be the case, it was decided to introduce locational data from a previous study that in addition to surveyed sites involved the analysis of the sites treated here @roalkvam2020. This found that the most consistently important locational variable for the sites in the region was the exposure of the sites to the elements, and failed to identify any diachronic variation in this pattern. As it has previously been proposed that sites located in areas exposed to wind and waves and with large overview of surrounding areas is a characteristic of hunting stations or similar, one could envisage that variablity in site functionality indicated by the core to chip correlation might be reflected in variation in degree of exposure.
